HealpixNet: Efficient spherical Convolutional Neural Networks with Healpix sampling for cosmological applications

### 2018-06-20 scnn.tex
done for today

Convolutional Neural Networks (CNNs) are becoming an important analysis tool in cosmology and astrophysics.
In many applications, the data comes in a form of spherical maps.
A popular format for these maps is Healpix, which uses equal-area, isolatitude sampling.
We present a CNNs for analysis of full- and partial sky Healpix maps, which we call HealpixCNN (???).
This approach is based on representing a sphere as a graph of connected nodes.
We describe a convolution operation defined on that graph.
HealpixCNN (???) uses Tensorflow as the underlying engine and can utilise most of it functionalities.
Architectures can be designed in a similar way as standard CNNs used for image analysis.
We find that the graph-based convolution is very efficient (MORE HERE)..
The filters learned by the HealpixCNN are restricted to being radial.
We demonstrate the method on a simple classification problem of mass maps from two cosmological models.
We compare the performance to a simple benchmark classifier, which uses histograms of map pixel values.
HealpixCNN achieves much better testing accuracy than the benchmark approach.
We make the code and examples available to the public.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

@@

\section{TODO list}
\begin{itemize}
    \item Michael, do you agree with the title HealpixNet
    \item Emphaise the novelty of using it on Helapix in abstract, into and rest of the paper
    \item Finish section on Convolutions on graphs (Michael - Nathanael)
    \item Plot of the new filters from Mass maps (Michael)
    \item (done!, plot to be added) check the power spectrum after smoothing
    \item (done!) upgrade the plot of pooling (Nathanael)
    \item new plot for network structure (teaser plot) (Michael)
    \item add info on comparison with histogram, so that it's just a demonstration
    \item (done!) Figure 1 to be round
    \item finish section on Efficient convolutions (Michael)
    \item conclusions
    \item (done!) new simulations with 3 arcmin smoothing
    \item (done!) simulation with cross-validation for the histogram
    \item Update result section according to the new simulations...
    \item (done!) simulation PSD
    \item (done!) Results figure, add sigma8
    \item possibly move the eigen decomposition and heat into appendix + add discussion (Nathanael)
\end{itemize}

@@

Formulations of deep learning on the sphere have been proposed by ~\citet{cohen2017convolutional,cohen2018spherical}.
In that method, a different sphere pixelisation scheme is used: the sphere pixels are mapped to flat space (??) and then generalized Fast Fourier Transform (FFT) is used for convolutions.
Other recently developed sampling methods include \citep{mcewen2011novelsampling}, where exact sampling theorem for equiangular MW sampling was presented.
In our work we focused on the Healpix sampling, as it is the most commonly used in cosmology.
It can be, however, extended to other sampling schemes, by straightforward change in definition of the graph.

Another advantage of our method is the speed of the convolution operation.
It is the most efficient spherical convolution\footnote{\todo{provably? cannot be faster than O(n) without approximations, e.g. sketching}}, requiring only $O(n)$ operations, where $n$ is the number of points.
MORE HERE

The method presented here restricts the shape of the filters learned by the network to be radial.
While this may be a less general design compared to the one used by conventional CNNs on flat images, we found it to still work efficiently for our test cases.
The layered structure of the network can partially compensate for that.
Further generalisations to non-radial filters may be possible [cite?], but we leave it to future work.

We give a practical demonstration of the application of our package to cosmological model discrimination using convergence maps, similar to \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify convergence maps on parts of a sphere into two cosmological models.
These models were designed to have the same power spectrum in range $\ell < 1000$.
We compare the performance of the spherical CNN to a baseline algorithm, a SVM classifier which takes pixel histograms of these maps as input.
The comparison is made as a function of noise level on convergence maps and area of the sphere used.

The paper is organised as follows.
In Section \ref{sec:related} we describe existing approaches to convolutions on the sphere, graphs, and manifolds.
Section \ref{sec:method} we show the way to construct a graph using Healpix sampling and define the convolution operation.
The comparison between the spherical CNN and the benchmark method using the weak lensing mass map classification is presented in Section \ref{sec:experiments}.
We conclude in Section \ref{sec:conclusion}.
% Cite the following papers:
% \begin{itemize}
%     \item Healpix paper \citep{gorski2005healpix}
%     \item First DES mass maps \citep{chang2017curvedsky}
%     \item Healpix convolutions with asymetric beams \citep{mitra2011fastpixel}
%     \item Mass mapping on the sphere \citep{wallis2017mappingdark}
%     \item Wavelets on the sphere \citep{leistedt2016wavelet} and ball \citep{leistedt2012exactwavelets}
%     \item Planck main result \citep{planck2015cosmologicalparameters,komatsu2011sevenyear}
%     \item HI map \citep{HI4PI2016fullskyHI,}
%     \item more radio maps from Adam+Hamsa \citep{santos2015cosmologySKA}
%     \item more astro science on the sphere, what other survey make maps?
%     \item alternative sampling on the sphere \citep{mcewen2011novelsampling} MW sampling
%     \item cite kids tomographic power spectrum \citep{koehlinger2017kidstomographic}
% \end{itemize}

% a) Cosmology has a lot of spherical data. Our method is simple and easy to use. Moreover, it is based on the widely used healpix sampling.
% b) It is the most efficient spherical convolution\footnote{\todo{provably? cannot be faster than O(n) without approximations, e.g. sketching}}, requiring only $O(n)$ operations, where $n$ is the number of points.

@@

% \section{Healpix sampling}
% Should we have this subsection to explain the advantages of Healpix?
% \begin{itemize}
% 	\item Define the spherical grid?
% 	\item reference the "cubed sphere"~\cite{ronchi1996cubed}?
% \end{itemize}

% \begin{figure}[!ht]
% \centering
% \includegraphics[width=0.45\textwidth]{figures/healpix-3layers.jpg}
% \caption{3 smaller sclales of the healpix sampling.}
% \label{fig:healpix_sampling}
% \end{figure}

@@

\subsection{Spherical convolutions}
The major challenge to define Convolution Neural Network (CNN) onto a
spherical domain is to handle a convolution that is suitable for this irregular sampling domain.
Three approaches exist to address this issue.

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2017convolutional,cohen2018spherical}. The goal of
these contributions it to use a rotational equivariant convolution on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e.
a projection on the spherical harmonics), a  multiplication in the spectral
domain and an inverse spherical Fourier transform. Hence the main
computational cost of a convolution comes from the two Fourier transforms.
Fortunately, the Fourier transform can be accelerated for special samplings
set (theoretically including HealPix, \cite{...}). The main advantage of this
approach is that it provides a mathematically well defined rotational
equivariant network. Nevertheless, even with these Fourier transform
acceleration, the convolutions remain expensive in comparison to a traditional
2-dimensional convolution, limiting the practical use of this approach. While
a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible. We also note that this approach cannot be trivially accelerated
if only one part of the sphere is used. This is impractical in the field of
cosmology as the data often observed only on a subpart of the sphere.

A second direction has been followed in~\cite{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the geodesic grid and the “cubed
sphere” grid defined by Ronchi et. al~\cite{ronchi1996cubed}. Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere.
% \nati{Michael: Under some very
% specific hypothesis, this second approach is a particular case of our method,
% i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
% on a grid == 2d convolution. Do you think we should mention that? I think we
% should not.}

@@

We are not the first one to leverage a graph to define a CNN on a sphere.
\cite{khasanova2017graph} use the same idea to handle omnidirectional images.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as \cite{khasanova2017graph} uses a
grid. Second, it uses a different graph CNN based on a different parametrisation
of the convolution kernel.

@@

\subsection{Convolutions on manifolds}
\todo{TK and NP: maybe we just describe convolutions on sphere and graph}

@@

\subsection{Convolutions on point clouds?}
\todo{TK and NP: maybe we just describe convolutions on sphere and graph}

@@

\begin{figure}[!ht]
\centering
% \includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\vspace{-0.5cm}
\includegraphics[width=0.45\textwidth]{figures/half_graph_4.pdf}
\vspace{-0.5cm}
\caption{
Top view of a Healpix graph for half the sphere using an nside parameter of $4$.
Nodes with 7 neighbors are highlighted.
}
\label{fig:healpix_graph_4}
\end{figure}

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{12 first graph Fourier eigenvectors.}
\label{fig:graph_harmonics}
\end{figure*}

As shown in Figure~\ref{fig:graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the HealPix sampling.

@@

The major difference with the classical convolution is that the convolution kernel
$h$ cannot be thought as a $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization purposes, one can
look at the effect of the convolution on a Kronecker, i.e. one column of the matrix
$h(\L)$. \nati{However, due to the non-regularity of the graph (i.e. the fact
that there is not perfect sampling on the sphere), this visualization will
differ from one node to another. In the specific case of the full sphere, these
differences are negligible in most of the cases. When considering only subpart of
the sphere, one will observe important border effects.}

\paragraph{Example: heat diffusion}
Let us consider the heat diffusion problem
\begin{equation} \label{eq:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\ref{eq:heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
are one generalization of the squared frequencies on the graph. Hence the kernel $K_t$ is also a generalization of the
Gaussian kernel on the sphere. In Figure~\ref{fig:gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. In  Figure~\ref{fig:gaussian_filters_visualization} of
Appendix~\ref{app:filter_visualization}, we show different visualizations of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(Figure~\ref{fig:gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.

@@

>>>>

Convolutional Neural Networks (CNNs) are becoming an important analysis tool in cosmology and astrophysics.
While CNN apply to images, in many applications, the data comes in a form of spherical maps with the popular Healpix format, which uses equal-area, isolatitude sampling.
Hence, we present a spherical CNNs for analysis of full- and partial sky Healpix maps, which we call HealpixCNN (???) \nati{HealpixNet??? To be decided}.
Our approach is based on representing a sphere as a graph of connected nodes.
We describe a convolution operation defined on that graph.
HealpixCNN (???) uses Tensorflow as the underlying engine and can utilize most of it functionalities.
Architectures can be designed in a similar way as standard CNNs used for image analysis.
We find that the graph-based convolution is very efficient (MORE HERE)..
The filters learned by the HealpixCNN are restricted to being radial.
We demonstrate the method on a simple classification problem of mass maps from two cosmological models.
We compare the performance to a simple benchmark classifier, which uses histograms of map pixel values.
HealpixCNN achieves much better testing accuracy than the benchmark approach.
We make the code and examples available to the public.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Spherical Convolutional Neural Network \sep SCNN \sep Healpix CNN \sep Graph CNN

@@

\section{TODO list}
\begin{itemize}
    \item Michael, do you agree with the title HealpixNet
    \item Emphasize the novelty of using it on Helapix in abstract, into and rest of the paper
    \item Finish section on Convolutions on graphs (Michael - Nathanael)
    \item Plot of the new filters from Mass maps (Michael)
    \item (done!, plot to be added) check the power spectrum after smoothing
    \item (done!) upgrade the plot of pooling (Nathanael)
    \item new plot for network structure (teaser plot) (Michael)
    \item add info on comparison with histogram, so that it's just a demonstration
    \item (done!) Figure 1 to be round
    \item finish section on Efficient convolutions (Michael)
    \item conclusions
    \item (done!) new simulations with 3 arcmin smoothing
    \item (done!) simulation with cross-validation for the histogram
    \item (done!) Update result section according to the new simulations...
    \item (done!) simulation PSD
    \item (done!) Results figure, add sigma8
    \item (more or less done!) possibly move the eigen decomposition and heat into appendix + add discussion (Nathanael I am not sure about the heat diffusion example...)
\end{itemize}

@@

Different formulations of deep learning on the sphere have been proposed in~\citet{cohen2017convolutional,cohen2018spherical,boomsma2017spherical,khasanova2017graph}.
In all of these methods, a different sphere pixelisation scheme is used, making them unsuitable for most cosmological application. 
Hence, our work we focused on the Healpix sampling, as it is the most commonly used in cosmology.
It can be, however, extended to other sampling schemes, by straightforward change in definition of the graph. These approach are discussed in Section~\ref{sec:spherical_conv}.
%\nati{I am not sure this goes there...} Other recently developed sampling methods include \citep{mcewen2011novelsampling}, where exact sampling theorem for equiangular MW sampling was presented.

An important advantage of our method is the speed of its convolution operation as it is of order $\mathcal{O}(n)$, where $n$ is the number of points, which is the lowest possible complexity for convolution without approximations, e.g. sketching.
While our implementation is slower than traditional convolution on a 2 dimensional image, it remain in the same order of magnitude.
MORE HERE

Because of the spherical domain, the method presented here restricts the shape of the filters learned by the network to be radial.
While this may be a less general design compared to the one used by conventional CNNs on flat images, we found it to still work efficiently for our test cases.
The layered structure of the network can partially compensate for that.
Further generalizations to non-radial filters may be possible [cite?], but we leave it to future work.

We give a practical demonstration of the application of our package to cosmological model discrimination using convergence maps, similar to \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify convergence maps on parts of a sphere into two cosmological models.
These models were designed to have the same power spectrum in range $\ell < 1000$.
We compare the performance of the spherical CNN to a baseline algorithm, an SVM classifier which takes pixel histograms or Power Spectrum Densities (PSD) of these maps as input.
The comparison is made as a function of noise level on convergence maps and area of the sphere used.

The paper is organised as follows.
In Section~\ref{sec:related}, we describe existing approaches to convolutions on the sphere, graphs, and manifolds.
In Section~\ref{sec:method}, we show the way to construct a graph using Healpix sampling and define the convolution operation.
The comparison between the spherical CNN and the benchmark method using the weak lensing mass map classification is presented in Section~\ref{sec:experiments}.
We conclude in Section~\ref{sec:conclusion}.

% Cite the following papers:
% \begin{itemize}
%     \item Healpix paper \citep{gorski2005healpix}
%     \item First DES mass maps \citep{chang2017curvedsky}
%     \item Healpix convolutions with asymetric beams \citep{mitra2011fastpixel}
%     \item Mass mapping on the sphere \citep{wallis2017mappingdark}
%     \item Wavelets on the sphere \citep{leistedt2016wavelet} and ball \citep{leistedt2012exactwavelets}
%     \item Planck main result \citep{planck2015cosmologicalparameters,komatsu2011sevenyear}
%     \item HI map \citep{HI4PI2016fullskyHI,}
%     \item more radio maps from Adam+Hamsa \citep{santos2015cosmologySKA}
%     \item more astro science on the sphere, what other survey make maps?
%     \item alternative sampling on the sphere \citep{mcewen2011novelsampling} MW sampling
%     \item cite kids tomographic power spectrum \citep{koehlinger2017kidstomographic}
% \end{itemize}

@@

% About the the different sphere sampling.
% \begin{itemize}
% 	\item Define the spherical grid?
% 	\item reference the "cubed sphere"~\cite{ronchi1996cubed}?
% \end{itemize}

@@

\subsection{Spherical convolutions}
\label{sec:spherical_conv}
The major challenge to define Convolution Neural Network (CNN) onto a
spherical domain is to handle a convolution that is suitable for this irregular sampling domain.
Three approaches exist to address this issue.

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2017convolutional,cohen2018spherical}. The goal of
these contributions it to use a rotational equivariant convolution on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e.
a projection on the spherical harmonics), a  multiplication in the spectral
domain and an inverse spherical Fourier transform. Hence the main
computational cost of a convolution comes from the two Fourier transforms.
Fortunately, the Fourier transform can be accelerated for special samplings
set (theoretically including HealPix, \cite{mohlenkamp1999fast,rokhlin2006fast}). 
The main advantage of this
approach is that it provides a mathematically well defined rotational
equivariant network. Nevertheless, even with these Fourier transform
acceleration, the convolutions remain expensive in comparison to a traditional
2-dimensional convolution, limiting the practical use of this approach. While
a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible. We also note that this approach cannot be trivially accelerated
if only one part of the sphere is used. This is impractical in the field of
cosmology as the data often observed only on a subpart of the sphere.

A second direction has been followed in~\cite{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the geodesic grid and the “cubed
sphere” grid defined by Ronchi et. al~\cite{ronchi1996cubed}. Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most computationally efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere.
% \nati{Michael: Under some very
% specific hypothesis, this second approach is a particular case of our method,
% i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
% on a grid == 2d convolution. Do you think we should mention that? I think we
% should not.}

@@

We are not the first one to leverage a graph to define a CNN on a sphere.
\cite{khasanova2017graph} use the same idea to handle omnidirectional images.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as~\cite{khasanova2017graph} uses a
grid. Second, it uses a different graph CNN based on a different parametrization
of the convolution kernel.

@@

\todo{Maybe we just describe convolutions on sphere and graph, i.e: do we really need this two subsections? Maybe simply merge them?}
\subsection{Convolutions on manifolds}

@@

\subsection{Convolutions on point clouds?}

@@

\begin{figure}[!ht]
\centering
% \includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\vspace{-0.5cm}
\includegraphics[width=0.35\textwidth]{figures/half_graph_4.pdf}
\vspace{-0.5cm}
\caption{
Top view of a Healpix graph for half the sphere using an nside parameter of $4$.
Nodes with 7 neighbors are highlighted. \label{fig:healpix_graph_4}
}

\end{figure}

@@

As shown in Figure~\ref{fig:graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the HealPix sampling. This topic is further discusses in Appendix~\ref{sec:spherical_harmonics}.

@@

The major difference with the classical convolution is that the convolution kernel
$h$ cannot be thought as a $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization purposes, one can
look at the effect of the convolution on a Kronecker, i.e. one column of the matrix
$h(\L)$. 
Due to the non-regularity of the graph (i.e. the fact
that there is not perfect sampling on the sphere), the convolution effect will slightly
differ from one node to another. In the specific case of the full sphere, these
differences are negligible in most of the cases. However, when considering only subpart of
the sphere, one can observe important border effects.

\paragraph{Example:}
Let us consider the heat diffusion problem
\begin{equation} \label{eq:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\ref{eq:heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
are one generalization of the squared frequencies on the graph. Hence the kernel $K_t$ is also a generalization of the
Gaussian kernel on the sphere. In Figure~\ref{fig:gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. In  Figure~\ref{fig:gaussian_filters_visualization} of
Appendix~\ref{app:filter_visualization}, we show different visualizations of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(Figure~\ref{fig:gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.

@@

\section{Spherical harmonics VS graph Fourier eigenvectors}
Figure~\ref{fig:graph_harmonics} displays the first 16 eigenvectors of the
graph Fourier basis used for the convolution. While their is a clear
resemblance with the spherical harmonic, the small irregularities in the
sampling (see~\ref{fig:healpix_graph_4}) have two important effects on the
graph Fourier eigenvectors.  First, some graph eigenvectors will be
localized~\cite{perraudin2018global}, i.e: they will span only a small subpart
of the sphere.  Second, as Nside increase toward infinity and hence the number
of pixel $n\rightarrow \infty$, we are still unsure if the graph eigenvectors
converge toward the spherical harmonics. We believe that it might be the case
for a different graph construction (fully connected) and that a proof could be
built on top of the work of~\cite{belkin2007convergence}. This is obviously
out of the scope of this contribution.

\label{sec:spherical_harmonics}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{12 first graph Fourier eigenvectors.}
\label{fig:graph_harmonics}
\end{figure*}
